An introduction to Docker for reproducible research, with examples from the R environment

### 2014-09-19 docker-paper/manuscript.Rmd
tweaking

Documentation on how to install and run code associated with published
research is another frequent barrier to replication.  @Lapp_2014 found
this impairs a researcher's ability to install and build the software
necessary, as even small holes in the documentation were found to be
major barriers, particularly for "novices" [Sensu @Garijo_2013] --
where novices may be experts in nearby languages but unfamiliar with
the package managers and other tools of the language involved). This
same problem is discussed in [@Clark_2014].  Imprecise documentation
goes well beyond issues of the software environment itself: incomplete
documentation of parameters involved meant as few as 30% of analyses
($n=34$) using the popular software STRUCTURE could be reproduced in the
study of [@Gilbert_2012].

@@

The problems highlighted here are not unique to _academic_ software,
but impact software development in general. While the academic research
literature has frequently focused on the development of workflow software
dedicated to particular domains, or otherwise to the use of virtual
machines, the software development community has recently emphasized a
philosophy (rather than a particular tool), known as _Development_ and
Systems _Operation_, or more frequently just "DevOps." The approach is
characterized by scripting, rather than documenting, a description of the
necessary dependencies for software to run, usually from the Operating
System (OS) on up.  @Clark_2014 describes the DevOps approach along
with both its relevance to reproducible research and examples of its
use in the academic research context. They identify the difficulties
I have discussed so far in terms of effective documentation:

@@

This elegantly addresses both the demand for simplicity of use (one
executes a script instead of manually managing the environmental setup)
and comprehensiveness of implementation. @Clark_2014 are careful to note
that this is not so much a technological shift as a philosophical one:

@@

Nevertheless, a growing suite of tools designed explicitly for this
purpose have rapidly replaced the use of general purpose tools (such
as Makefiles, bash scripts) to become synonymous with the DevOps
philosophy.  @Clark_2014 reviews many of these DevOps tools, their
different roles, and their application in reproducible research.

I focus the remainder of this paper on one of the most recent and
rapidly growing among these, called Docker, and the role it can play
in reproducible research. Docker offers several promising features
for reproducibility that go beyond the tools highlighted in @Clark_2014.
Nevertheless, my goal in focusing on this technology is not to
promote a particular solution, but to anchor the discussion of technical
solutions to reproducibility challenges in concrete examples.

@@

Sharing these images is further facilitated by the Docker Hub technology. 
While Docker images tend to be much smaller than equivalent virtual
machines, moving around even 100's of gigabytes can be a challenge.
In their work on virtual machines, @Dudley_2010 recommend only running
these tools on cloud servers such as the Amazon EC2 system. As most
researchers still develop software locally and may not have ready access
to these resources, such a requirement adds an additional barrier to
reuse.  Docker provides a convenient way to share any image publicly
or privately through the Hub after creating a free account.  Docker Hub
is both a publicly available service and also available as 
a separate open source platform that can be deployed on private servers.

>>>>

Documentation on how to install and run code associated with published
research is another frequent barrier to replication. A study by Lapp [@Lapp_2014] found
this impairs a researcher's ability to install and build the software
necessary, as even small holes in the documentation were found to be
major barriers, particularly for "novices" [Sensu @Garijo_2013] --
where novices may be experts in nearby languages but unfamiliar with
the package managers and other tools of the language involved). This
same problem is discussed in [@Clark_2014].  Imprecise documentation
goes well beyond issues of the software environment itself: incomplete
documentation of parameters involved meant as few as 30% of analyses
($n=34$) using the popular software STRUCTURE could be reproduced in the
study of [@Gilbert_2012].

@@

The problems highlighted here are not unique to _academic_ software,
but impact software development in general. While the academic research
literature has frequently focused on the development of workflow software
dedicated to particular domains, or otherwise to the use of virtual
machines, the software development community has recently emphasized a
philosophy (rather than a particular tool), known as _Development_ and
Systems _Operation_, or more frequently just "DevOps." The approach is
characterized by scripting, rather than documenting, a description of the
necessary dependencies for software to run, usually from the Operating
System (OS) on up. Clark _et al._ [@Clark_2014] describe the DevOps approach along
with both its relevance to reproducible research and examples of its
use in the academic research context. They identify the difficulties
I have discussed so far in terms of effective documentation:

@@

This elegantly addresses both the demand for simplicity of use (one
executes a script instead of manually managing the environmental setup)
and comprehensiveness of implementation. Clark _et al._ [@Clark_2014] are careful to note
that this is not so much a technological shift as a philosophical one:

@@

Nevertheless, a growing suite of tools designed explicitly for this
purpose have rapidly replaced the use of general purpose tools (such
as Makefiles, bash scripts) to become synonymous with the DevOps
philosophy.  Clark _et al._ [@Clark_2014] reviews many of these DevOps tools, their
different roles, and their application in reproducible research.

I focus the remainder of this paper on one of the most recent and
rapidly growing among these, called Docker, and the role it can play
in reproducible research. Docker offers several promising features
for reproducibility that go beyond the tools highlighted in [@Clark_2014].
Nevertheless, my goal in focusing on this technology is not to
promote a particular solution, but to anchor the discussion of technical
solutions to reproducibility challenges in concrete examples.

@@

Sharing these images is further facilitated by the Docker Hub technology. 
While Docker images tend to be much smaller than equivalent virtual
machines, moving around even 100's of gigabytes can be a challenge.
In their work on virtual machines, Dudley _et al._ [@Dudley_2010] recommend only running
these tools on cloud servers such as the Amazon EC2 system. As most
researchers still develop software locally and may not have ready access
to these resources, such a requirement adds an additional barrier to
reuse.  Docker provides a convenient way to share any image publicly
or privately through the Hub after creating a free account.  Docker Hub
is both a publicly available service and also available as 
a separate open source platform that can be deployed on private servers.
